---
title: "Lab 2"
author: 'Jiajia Feng'
date: ' `r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
```


## Question 1
### b
```{r}
n <- 200
delta1 <- 0.6
delta2 <- 0.4

set.seed(123)
Z <- rnorm(n)

# Simulate random walks
X1 <- cumsum(Z) + delta1 * (1:n)
X2 <- cumsum(Z) + delta2 * (1:n)

data <- data.frame(
  time = 1:n,
  RandomWalk1 = X1,
  RandomWalk2 = X2
)

ggplot(data, aes(x = time)) +
  geom_line(aes(y = RandomWalk1, colour = "Random Walk 1 (delta = 0.6)")) +
  geom_line(aes(y = RandomWalk2, colour = "Random Walk 2 (delta = 0.4)")) +
  labs(title = "Random Walk Simulations",
       x = "Time",
       y = "Value") +
  scale_colour_manual("", 
                      breaks = c("Random Walk 1 (delta = 0.6)", "Random Walk 2 (delta = 0.4)"),
                      values = c("blue", "red"))


```
### c
```{r}
"Both random walks should show a clear upward trend over time. 
The series with delta = 0.6 will have a steeper slope compared to the series with delta = 0.4."
"Around these two trend, fluctuations are random and centered around the trend line dictated by 
the respective delta value."
#non-stationary
"The mean of the series increases over time due to the presence of the delta term.  
This is visible as a consistent upward trend in the plot.
The variance also increases over time because the effect of the white noise accumulates."
"The random walk is inherently non-stationary.
The mean is not constant over time due to the drift component (delta), 
and the variance is not constant because the effects of the shocks accumulate over time."
```

## Question 2
### b
```{r}
library(forecast)

phi1 <- 0.35
phi2 <- 0.45
ar_coefs <- c(phi1, phi2)

n <- 200
set.seed(123)
ar2_process <- arima.sim(n = n, model = list(ar = ar_coefs), sd = 1)


plot(ar2_process, main = "AR(2) Process Simulation", ylab = "X_t", xlab = "Time")

acf(ar2_process, lag.max = 20, main = "ACF of AR(2) Process")
pacf(ar2_process, lag.max = 20, main = "PACF of AR(2) Process")

```

### c
```{r}
theta1 <- 0.45
theta2 <- 0.55
ma_coefs <- c(theta1, theta2)

n <- 200
set.seed(123)
ma2_process <- arima.sim(n = n, model = list(ma = ma_coefs), sd = 1)

theoretical_acf <- c(1, (theta1 + theta1*theta2)/(1 + theta1^2 + theta2^2), 
                     theta2/(1 + theta1^2 + theta2^2))

#Sample ACF
acf_values <- acf(ma2_process, lag.max = 20, plot = FALSE)

# Combined plot
plot(acf_values, main = "Sample and Theoretical ACF for MA(2) Process", xlab = "Lag", ylab = "ACF")
points(0:2, theoretical_acf, col = "blue", pch = 20)

# Sample PACF
pacf(ma2_process, lag.max = 20, main = "Sample PACF for MA(2) Process")

```
### d
```{r}
set.seed(123)

#AR(2) process
ar_coefs <- c(0.35, 0.45)
ar2_process <- arima.sim(n = 200, model = list(ar = ar_coefs), sd = 1)

#MA(2) process
ma_coefs <- c(0.45, 0.55)
ma2_process <- arima.sim(n = 200, model = list(ma = ma_coefs), sd = 1)

# Plot ACF for AR(2) process
acf_ar2 <- acf(ar2_process, lag.max = 20, plot = FALSE)

# Plot ACF for MA(2) process
acf_ma2 <- acf(ma2_process, lag.max = 20, plot = FALSE)

#Combined ACF plot
par(mfrow=c(2, 1))
plot(acf_ar2, main = "Sample ACF for AR(2) Process")
plot(acf_ma2, main = "Sample ACF for MA(2) Process")

# Plot PACF for AR(2) process
pacf_ar2 <- pacf(ar2_process, lag.max = 20, plot = FALSE)

# Plot PACF for MA(2) process
pacf_ma2 <- pacf(ma2_process, lag.max = 20, plot = FALSE)

# Combined PACF plot
par(mfrow=c(2, 1))
plot(pacf_ar2, main = "Sample PACF for AR(2) Process")
plot(pacf_ma2, main = "Sample PACF for MA(2) Process")

```

Comparison
"Sample ACF Comparison: 
AR(2) Process: 
The sample ACF of an AR(2) process shows a gradual decay in correlation as the lags increase.
MA(2) Process: 
In contrast, the sample ACF of an MA(2) process show significant correlations at the first two lags 
and then drop to zero or non-significant values for lags beyond the order of the MA process 
(after lag 2 in this case)."

"Sample PACF Comparison:
AR(2) Process: 
The sample PACF of an AR(2) process show significant correlations at the first two lags, 
which corresponds to the order of the process. After the first two lags, the PACF drop off.
MA(2) Process: 
The sample PACF for an MA(2) process doesn't have a sharp cutoff like the ACF. 
It shows a more complex and slower decay pattern as the PACF."

## Question 3
```{r}
wine.csv <- read.table("monthly-australian-wine-sales-th.csv", 
                       sep = ",", header = FALSE, skip = 1, nrows = 187)
colnames(wine.csv) <- c("Month", "Sales")
wine = ts(wine.csv$Sales, start = c(1980, 1), frequency = 12) 
ts.plot(wine, main = "Wine Sales over Time")
```

### a
```{r}
wine_log <- log(wine.csv$Sales)

# Plot the transformed data
plot.ts(wine_log, main = "Log-transformed Wine Sales over Time")

bc_transform <- BoxCox.lambda(wine.csv$Sales)
wine_boxcox <- BoxCox(wine.csv$Sales, bc_transform)

# Plot the Box-Cox transformed data
plot.ts(wine_boxcox, main = "Box-Cox Transformed Wine Sales over Time")

```

### b
```{r}

wine_ts <- ts(wine.csv$Sales, start = c(1980, 1), frequency = 12)

diff_wine_ts <- diff(wine_ts, differences = 1)

plot.ts(diff_wine_ts, main = "Differenced Wine Sales over Time", 
        ylab = "Differenced Sales", xlab = "Time")

abline(h = 0, col = "red")

```
### c
```{r}

Acf(diff_wine_ts, lag.max = 60, main = "ACF of Differenced Wine Sales")

Pacf(diff_wine_ts, lag.max = 60, main = "PACF of Differenced Wine Sales")
```

ACF Plot: 
If there is still cyclic behavior or seasonality, the ACF plot shows a pattern of spikes 
at regular intervals.

PACF Plot:
The presence of significant lags in the PACF plot suggests additional AR or seasonal AR terms 
are needed in the model.

The reason we can still see a repeating pattern in the ACF plot, even after differencing,
due to seasonality in the data that has not been addressed by simply differencing at lag 1.


### d
```{r}

seasonally_diff_wine_ts <- diff(diff_wine_ts, lag = 12)

plot.ts(seasonally_diff_wine_ts, main = "Seasonally Differenced Wine Sales over Time", 
        ylab = "Seasonally Differenced Sales", xlab = "Time")


abline(h = 0, col = "red")

Acf(seasonally_diff_wine_ts, lag.max = 60, main = "ACF of Seasonally Differenced Wine Sales")
Pacf(seasonally_diff_wine_ts, lag.max = 60, main = "PACF of Seasonally Differenced Wine Sales")

```

